---
title: "Assignment: China Temperature Analysis"
subtitle: "Geospatial Analysis of Climate Change in China"
author: "Your Name Here"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float: true
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.width = 10, fig.height = 6)
```

## Instructions

In this assignment, you will analyze temperature data for China using the *same techniques* we learned with precipitation data. You have been provided with two temperature raster files:

- `china_historic_temp.tiff` - Historic mean annual temperature (1970-2000) in degrees Celsius
- `china_2050_temp.tiff` - Projected mean annual temperature (2041-2060) in degrees Celsius

Complete the following tasks using R and the geospatial packages we learned about in class.

---

## Setup

Make sure you have loaded all necessary packages:

```{r load-packages}
library(tidyverse)
library(sf)
library(terra)
library(tmap)
library(rnaturalearth)
```

Load the spatial objects we created in class:

```{r load-spatial-data}
# Load China boundary data
china_mainland <- ne_countries(country = "China", returnclass = "sf", scale = "medium")
china_provinces <- ne_states(country = "China", returnclass = "sf")

# Create cities dataset
chinese_cities <- data.frame(
  city_name = c("Beijing", "Shanghai", "Guangzhou", "Shenzhen", 
                "Chengdu", "Hangzhou", "Xi'an", "Wuhan", 
                "Nanjing", "Tianjin"),
  longitude = c(116.4074, 121.4737, 113.2644, 114.0579,
                104.0668, 120.1551, 108.9480, 114.3054,
                118.7969, 117.1901),
  latitude = c(39.9042, 31.2304, 23.1291, 22.5431,
               30.6728, 30.2741, 34.3416, 30.5928,
               32.0603, 39.1255),
  population_millions = c(21.9, 24.9, 18.8, 17.8,
                         21.4, 8.6, 9.22, 8.99,
                         10.17, 14.7)
)

# Convert to spatial data
cities_sf <- st_as_sf(chinese_cities, 
                      coords = c("longitude", "latitude"),
                      crs = 4326)
```

---

## Question 1: Load Temperature Raster Data (5 points)

Load both temperature raster files using the `rast()` function:

```{r}
# Load temperature raster data
china_temp_hist <- # YOUR CODE HERE
china_temp_2050 <- # YOUR CODE HERE

```


---

## Question 2: Visualize Historic Temperature (10 points)

Create a map showing historic temperature across China. Your map should:
- Convert the raster to a data frame for plotting
- Use `geom_raster()` to display temperature values
- Add the mainland China country boundary as an overlay
- Include appropriate color scale and labels
- Use `scale_fill_distiller()` and the "RdYlBu" palette to make the colors more appropriate for viewing temperature data.

```{r}
# YOUR CODE HERE
```


---

## Question 3: Calculate Temperature Change (10 points)

Calculate the projected change in temperature between 2050 and historic. Call the new raster object "temp_change" and use `plot()` to view it.

```{r}
# YOUR CODE HERE
```

---

## Question 4: Extract Temperature Data for Cities (15 points)

Extract the projected temperature change values for all major Chinese cities using the `extract()` function and add this information as a new column in the `cities_sf` dataset:

```{r}
# YOUR CODE HERE
```


Create a map showing cities sized by their projected temperature change:

```{r}
# YOUR CODE HERE
```



Which cities are projected to experience the greatest temperature increases? List the top 3 and their approximate temperature changes.

*Your answer here:*

---

## Question 5: Provincial Temperature Analysis (15 points)

Calculate the mean projected temperature change for each province using the `zonal()` function. Add the values as a new column to your china_provices data.

```{r}
# YOUR CODE HERE
```



Create a map showing provinces colored by their projected temperature change:

```{r}
# YOUR CODE HERE
```



Which provinces are projected to warm the most and least?

*Your answer here:*


---

## Bonus Task: Change Correlations (5 extra credit points)

Join the *provinces_precip* dataset we created togther in class with the similar *provinces_temp* dataset (or whatever you named it) you created in this assignment and create a scatter plot that shows the relationship between change in precipitation and change in temperature. What trends do you see?

Notes:
- To use any of the dplyr join functions (such as `left_join()`) you will first need to convert *provinces_precip* and your new *provinces_temp* object from an sf object to a data frame. You can do this with the `st_drop_geometry()` function.

```{r bonus-task}
# Your bonus code here (if attempting)

```



