---
title: "ASSIGNMENT 5: Data Wrangling"
author: "Katie Willi"
date: "2025-06-08"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Instructions

Complete all 5 questions below using the National Park Service visitation datasets. Make sure to load the necessary packages and data at the beginning of your assignment. Each question builds upon concepts covered in our functions, iteration, and data wrangling lessons.

```{r}
# Load packages
library(tidyverse)
library(plotly)
library(scales)

# Read in the data
parkwide_data <- read_csv("data/nps_parkwide_visitation.csv")
unit_data <- read_csv("data/nps_unit_visitation.csv")

# Set up the data as shown in class
years <- (1980:2024)
parks <- c("ROMO", "ACAD", "LAKE", "YELL", "GRCA", "ZION", "OLYM", "GRSM")

parkwide <- parkwide_data %>%
  filter(Year %in% years)

units <- unit_data %>% 
  filter(Year %in% years & UnitCode %in% parks)

# Combine datasets
visitation <- bind_rows(parkwide, units)

# Create annual visitation data
annual_visitation <- visitation %>%
  group_by(UnitCode, Year) %>% 
  summarize(RecVisitation = sum(RecreationVisitors))
```

#### Question 1: Calculate Percentage of Total Park Visitation 

Task: Calculate a new variable showing each park's *percentage* of RecVisitation compared to the total RecVisitation.

Create a new dataframe called `park_percentages` with columns: Year, UnitCode, RecVisitation, and PercentOfTotal.

Hint: Use `group_by(Year)` then `mutate()` to calculate percentages within each year.

```{r}
# YOUR CODE HERE #
```

#### Question 2a: Create a Plotting Function 

Write a function called plot_monthly_visitation() that takes one argument:

`park_code`: a single park unit code (example: "YELL")

The function should include three steps:
1) Filter the `units` dataframe to the selected park (using the `park_code` argument)
2) Using the filtered data, calculate the average monthly RecreationVisitors (using `group_by()` and `summarise()`)
3) Create a bar plot that shows those calculated monthly averages for the selected park (using `geom_col()` with the month on the x axis and the average visitation on the y axis)

The function should return a bar plot showing the average monthly recreational visitation for the selected park across all available years. The plot should have proper labels and formatting.

**Then test your function using "YELL" (Yellowstone) for the `park_code = ` argument.**


```{r}
# YOUR CODE HERE #
```

#### Question 2b: 

Use `map()` to create plots for multiple parks using the function you made in Question 2a and the list of park names below (`park_list`).

```{r}
park_list <- c("YELL", "GRCA", "ZION")
```


```{r}
# YOUR CODE HERE #
```

#### Question 3: Understanding Different Join Types

**Execute the section of code below, inspect the results, and then answer the questions about the different join functions below.**

This code creates two smaller datasets:

niob_data: Filter unit_data for Niobrara National Scenic River (UnitCode = "NIOB") for years 2020-2024
parkwide_subset: Filter parkwide_data for years 2023-2024

And then performs each of the following joins and examines the results:

```{r}
# Create the datasets
niob_data <- unit_data %>% 
  filter(UnitCode == "NIOB", Year %in% 2020:2024)

parkwide_subset <- parkwide_data %>% 
  filter(Year %in% 2023:2024)

# Perform each join and examine the number of rows returned
inner_result <- inner_join(niob_data, parkwide_subset, by = c("Year", "Month"))
left_result <- left_join(niob_data, parkwide_subset, by = c("Year", "Month"))
right_result <- right_join(niob_data, parkwide_subset, by = c("Year", "Month"))
anti_result <- anti_join(niob_data, parkwide_subset, by = c("Year", "Month"))

# Check the number of rows in each result
nrow(inner_result)
nrow(left_result)
nrow(right_result)
nrow(anti_result)

# View the output of each result
```

**ANSWER THE QUESTIONS BELOW**: 

1) What is the difference between `inner_join()` and `anti_join()`?

2) What is the difference between `left_join()` and `right_join()`?


#### Question 4: Peak Season Analysis with Joins 

Identify each park's busiest month in 2023, then join this information with the park's total annual visitation for 2023.

**Steps**:

1. Find the month that has the maximum visitors for each park in 2023 (*we did this part for you*):
```{r}
# Step 1: Find peak month for each park in 2023
peak_months <- units %>%
  filter(Year == 2023) %>%
  group_by(UnitCode) %>%
  slice_max(RecreationVisitors, n = 1) %>%
  select(UnitCode, MaxMonth = Month, MaxMonthVisitation = RecreationVisitors)
```

2. Using `peak_months` created above, calculate total visitation for each park in 2023 (using `filter()`, `group_by()` and then `summarise()`)
```{r}
#YOUR CODE HERE
```

3. Join the dataframes created in Step 1 and Step 2 together using the "UnitCode" variable for the `by = ` argument.

*Your final result should show: UnitCode, MaxMonth, MaxMonthVisitation, and TotalAnnualVisitation for 2023.*

```{r}
# YOUR CODE HERE #
```


#### Question 5: Multi-Year Comparison with Advanced Joining 

Compare how each park's 2020 visitation (pandemic year) compared to their 2019 and 2021 visitation using joins. Create a final dataset that shows for each park:

Create final result showing:
- 2019, 2020, 2021 total visitation
- Percent change from 2019→2020  
- Percent change from 2020→2021

Each park is a row in the data frame, and each variable above is a column. 

**STEPS**

1. Get annual visitation for 2019, 2020 and 2021
```{r}
three_year_data <- annual_visitation %>%
  filter(Year %in% c(2019, 2020, 2021)) %>%
  select(UnitCode, Year, RecVisitation)
```

2. Use `pivot_wider()` to make each Year its own column
```{r}
# YOUR CODE HERE
```

3. Using the "wide" data created above, calculate two new columns that represent the change in visitation from Year 2019 to 2020 and from Year 2020 to 2021.

**Hint**: Use the formula: ((new_value - old_value) / old_value) * 100 for percent change.

```{r}
# YOUR CODE HERE 
```

