---
title: "ASSIGNMENT_1_Intro"
output: word_document
---

To answer each question, put the code in a code chunk directly below each question. For this assignment you will need to load the `tidyverse` and `palmerpenguins` packages, and use the `penguins` dataset like we did in the lesson.

A reminder, you can load in the `penguins` data with the line of code below:
```{r}
data("penguins")
```


#### Exercises


1.  Why don't the following lines of code work? Fix each one so the code runs

```{r eval=FALSE}
myList["Fruit"]
```

```{r eval=FALSE}
penguins$flipper_lenght_mm
```

```{r eval=FALSE}
penguins[island=='Dream',]
```



2.  How many species are in the `penguins` data set? What islands were the data collected for? (Note: the `unique()` function might help)




3.  Use indexing to create a new data frame that has only 3 columns: species, island and flipper length columns, and subset all rows for just the 'Dream' island.





4.  Use indexing and the `mean()` function to find the average flipper length for the *Adelie* species on Dream island. (Note: explore the `mean()` function and how to deal with NA values).





5.  Reorder the variables in `penguins` so that `year` is the first column followed by the rest (Hint: look into the use of `everything()`).





6.  Create a new column called 'size_group' where individuals with body mass greater than 4000 are called 'large' and those smaller are called 'small'.






7.  Which species had the largest average body mass. Remember to account for NA values.






8.  Filter data for years that are *not* 2008 or 2009.





