---
title: "ASSIGNMENT_1_Intro"
output: word_document
---

To answer each question, put the code in a code chunk directly below each question. For this assignment you will need to load the `tidyverse` and `palmerpenguins` packages, and use the `penguins` dataset like we did in the lesson.

A reminder, you can load in the `penguins` data with the line of code below:
```{r}
data("penguins")
```


#### Exercises


1.  Why don't the following lines of code work? Fix each one so the code runs.

```{r eval=FALSE}
myList["Fruit"]
```

```{r eval=FALSE}
penguins$flipper_lenght_mm
```

```{r eval=FALSE}
penguins[island=='Dream',]
```

**ANSWERS**
```{r}
myList[["Fruit"]] #double brackets when indexing a list

penguins$flipper_length_mm # "length" was spelled wrong

penguins[penguins$island == 'Dream',] # needed to specify the data frame that "island" was a variable in
```


2.  How many species are in the `penguins` data set? What islands were the data collected for? Note: the `unique()` function might help.

**Answer**

```{r}
unique(penguins$species) # Adelie, Gentoo, Chinstrap. 3 species total

unique(penguins$island) # Torgersen, Biscoe and Dream island.
```



3.  Use indexing to create a new data frame that has only 3 columns: species, island and flipper length columns, and subset all rows for just the 'Dream' island.

**Answer**

```{r}
#indexing
penguins[penguins$island == "Dream", c("species", "island", "flipper_length_mm")]

# OR, you could do this with the `select()` and `filter()` functions
penguins %>% 
  filter(island == "Dream") %>% 
  select(species, island, flipper_length_mm)
```



4.  Use indexing and the `mean()` function to find the average flipper length for the *Adelie* species on Dream island. Note: explore the `mean()` function and how to deal with NA values.

**Answer**

```{r}
values <- penguins[penguins$species == "Adelie" & penguins$island == "Dream", "flipper_length_mm"]

# to get the mean:
mean(values$flipper_length_mm, na.rm = TRUE) # 189.7321

#OR, using dplyr functions:
penguins %>% 
  filter(species == "Adelie" & island == "Dream") %>% 
  summarise(mean = mean(flipper_length_mm, na.rm = TRUE)) # Rounds the value to 190
```



5.  Reorder the variables in `penguins` so that `year` is the first column followed by the rest. Hint: look into the use of `everything()`.

**Answer**

```{r}
penguins %>% 
  select(year, everything())

```



6.  Create a new column called 'size_group' where individuals with a body mass greater than 4000 are called 'large' and those smaller are called 'small'.


**Answer**

```{r}
penguins %>% 
  mutate(size_group = if_else(body_mass_g > 4000, "large", "small"))
```



7.  Which species had the largest average body mass? Remember to account for NA values.


```{r}
penguins %>% 
  group_by(species) %>% 
  summarise(avg_body_mass = mean(body_mass_g, na.rm = TRUE))

# Gentoo are the largest
```



8.  Filter data for years that are *not* 2008 or 2009.


```{r}
penguins %>% 
  filter(!year %in% c("2008", "2009"))
```



