---
title: "Data Exploration Assignment: Bison Dataset"
author: "Katie Willi"
date: "2025-06-03"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

## Introduction

In this assignment, we will be using a new dataset containing information collected about bison on the Konza Prairie in Kansas, USA.

**Data Credit:**
EBlair, J. 2021. CBH01 Konza Prairie bison herd information ver 12. Environmental Data Initiative. https://doi.org/10.6073/pasta/9c641b35695abc5889edd64c3950517f (Accessed 2021-05-10). doi:10.6073/pasta/9c641b35695abc5889edd64c3950517f

**Dataset Variables:**
- `data_code`: a character denoting the dataset code
- `rec_year`: a number denoting the year of observation
- `rec_month`: a number denoting the month of observation
- `rec_day`: a number denoting the day of observation
- `animal_code`: a character denoting the unique individual bison identification code based on ear tag number
- `animal_sex`: a character denoting the sex of bison: M = male, F = female, U = unknown
- `animal_weight`: a number denoting bison weight in pounds
- `animal_yob`: a number denoting the year animal was born

## Setup

First, load the necessary packages and read in the bison dataset:

```{r}
library(tidyverse)

# Read in the bison dataset
bison <- read_csv("data/bison_data.csv")
```

## Exercise 1: Basic Data Exploration

Use the following functions to explore the bison dataset:

a) Use `nrow()` and `ncol()` to find how many rows and columns the dataset has
b) Use `names()` to see the column names
c) Use `summary()` to get a basic summary of all variables

```{r}
# Your code here

```

## Exercise 2: Simple Filtering

Filter the bison dataset to include only female bison ("F"). 
Store this in a new object called `female_bison`.

```{r}
# Your code here

```

## Exercise 3: Creating a New Variable

Use `mutate()` to create a new column called `age_at_recording` that calculates how old each bison was when recorded.

```{r}
# Your code here

```

## Exercise 4: Using if_else() for Categories

Create a new column called `weight_category` using `if_else()`. Classify bison as:
- "Heavy" if their weight is greater than 800 pounds
- "Light" if their weight is 800 pounds or less

```{r}
# Your code here

```

## Exercise 5: Basic Grouping and Summarizing

Find the average weight for each sex (M, F, U). 

```{r}
# Your code here

```

## Exercise 6: Selecting and Arranging

Select only the columns: animal_code, animal_sex, animal_weight, and rec_year.
Then arrange the data by weight from heaviest to lightest using `arrange()`.

Hint: Look at the examples in `?arrange`.

```{r}
# Your code here

```

## Exercise 7: Basic Scatter Plot

Create a scatter plot using `geom_point()` to show the relationship between recording year (x-axis) and bison weight (y-axis):
- Color the points by sex 
- Add appropriate axis labels

```{r}
# Your code here

```

## Bonus Question

Look at your scatter plot from Exercise 8. Do you notice any patterns? Write 1-2 sentences about what you observe.

**Your observation:**
[Write what you notice here]
