---
title: "Joins and Pivots in R"
date: "2025-06-09"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

## Part 1: Introduction to Joins

### What are joins?

Joins combine data from two datasets based on common columns (keys).

**Real-world example**: You have student names and IDs in one file, and their test scores in another file. You want to combine them to see which student got which score.

```{r}
library(tidyverse)

# Create example datasets
students <- data.frame(
  student_id = c(1, 2, 3, 4),
  name = c("Alice", "Bob", "Charlie", "Diana"),
  major = c("Math", "English", "Math", "Biology")
)

test_scores <- data.frame(
  student_id = c(1, 2, 3, 5),  # Note: student 4 missing, student 5 extra
  test_score = c(85, 92, 78, 88),
  test_date = c("2023-01-15", "2023-01-15", "2023-01-16", "2023-01-16")
)

print("Students table:")
students

print("Test scores table:")
test_scores
```

------------------------------------------------------------------------

## Part 2: The Five Main Join Types

### 1. Inner Join - Keep only matching rows

**Rule**: Only keep rows that exist in BOTH datasets

```{r}
inner_result <- inner_join(students, test_scores, by = "student_id")
print("Inner Join Result:")
inner_result
```

**What happened?**

\- Only students 1, 2, 3 appear (they exist in both datasets)

\- Student 4 (Diana) disappeared - no test score

\- Student 5's score disappeared - no student info

------------------------------------------------------------------------

### 2. Left Join - Keep all from left dataset

**Rule**: Keep ALL rows from the left (first) dataset, add matching data from right

```{r}
left_result <- left_join(students, test_scores, by = "student_id")
print("Left Join Result:")
left_result
```

**What happened?**

\- ALL students kept (1, 2, 3, 4)

\- Diana (student 4) has NA for test_score and test_date

\- Student 5's score still disappeared

------------------------------------------------------------------------

### 3. Right Join - Keep all from right dataset

**Rule**: Keep ALL rows from the right (second) dataset, add matching data from left

```{r}
right_result <- right_join(students, test_scores, by = "student_id")
print("Right Join Result:")
right_result
```

**What happened?**

\- ALL test scores kept (students 1, 2, 3, 5)

\- Student 5 has NA for name and major

\- Diana (student 4) disappeared

------------------------------------------------------------------------

### 4. Full Join - Keep everything

**Rule**: Keep ALL rows from BOTH datasets

```{r}
full_result <- full_join(students, test_scores, by = "student_id")
print("Full Join Result:")
full_result
```

**What happened?**

\- Everyone appears (students 1, 2, 3, 4, 5)

\- Missing data filled with NA

------------------------------------------------------------------------

### 5. Anti Join - Keep non-matching rows from left

**Rule**: Keep rows from the left dataset that do NOT have matches in the right dataset

```{r}
anti_result <- anti_join(students, test_scores, by = "student_id")
print("Anti Join Result:")
anti_result
```

**What happened?**

\- Only Diana (student 4) appears

\- She's the only student who did NOT take the test

\- No columns from test_scores are included

**When is this useful?**

\- Find students who missed the test

 - Identify customers who haven't made purchases

\- Find data quality issues (unmatched records)

------------------------------------------------------------------------

## Part 3: Quick Reference - When to Use Which Join

| Join Type | When to Use | Example |
|------------------------|----------------------------|--------------------|
| `inner_join()` | Only want complete cases | "Show only students who took the test" |
| `left_join()` | Keep all from main dataset | "Show all students, even those who missed the test" |
| `right_join()` | Keep all from second dataset | "Show all test scores, even mystery students" |
| `full_join()` | Don't want to lose any data | "Keep everything for investigation" |
| `anti_join()` | Find what's missing | "Show students who DIDN'T take the test" |

**Most common**: `left_join()` - usually you have a main dataset and want to add information to it.

------------------------------------------------------------------------

## Part 4: Introduction to Pivots

### What are pivots?

Pivots reshape data between "wide" and "long" formats.

**Wide format**: Each variable is a column **Long format**: Multiple observations per row, with variables in rows

```{r}
# Create example: student grades in different subjects
grades_wide <- data.frame(
  student = c("Alice", "Bob", "Charlie"),
  math = c(85, 92, 78),
  english = c(88, 85, 92),
  science = c(90, 88, 85)
)

print("Wide format:")
grades_wide
```

------------------------------------------------------------------------

### pivot_longer() - Wide to Long

**Use when**: You have multiple columns that represent the same type of measurement

```{r}
# Convert to long format
grades_long <- grades_wide %>%
  pivot_longer(cols = c(math, english, science),  # columns to pivot
               names_to = "subject",              # name for the new column
               values_to = "grade")               # name for the values

print("Long format:")
grades_long
```

**Why use long format?** - Easier to make plots with ggplot2 - Better for statistical analysis - More flexible for grouping and summarizing

------------------------------------------------------------------------

### pivot_wider() - Long to Wide

**Use when**: You want to compare values side by side

```{r}
# Convert back to wide format
grades_wide_again <- grades_long %>%
  pivot_wider(names_from = subject,    # column that becomes new column names
              values_from = grade)     # column that provides the values

print("Back to wide format:")
grades_wide_again
```

------------------------------------------------------------------------

## Part 5: Practical Examples

### Example 1: Combining Student Data

```{r}
# Real scenario: student info + course enrollments
student_info <- data.frame(
  id = 1:4,
  name = c("Amy", "Ben", "Carl", "Dana"),
  year = c(2, 1, 3, 2)
)

course_grades <- data.frame(
  id = c(1, 1, 2, 3, 3, 4),
  course = c("Math101", "Bio201", "Math101", "Eng150", "Bio201", "Eng150"),
  grade = c("A", "B+", "B", "A-", "B", "C+")
)

# Join to see student names with their grades
student_grades <- left_join(student_info, course_grades, by = "id")
print("Student grades:")
student_grades
```

### Example 2: Reshape for Analysis

```{r}
# Convert to wide format to compare courses
grades_comparison <- student_grades %>%
  select(name, course, grade) %>%
  pivot_wider(names_from = course, values_from = grade)

print("Grades comparison:")
grades_comparison
```

------------------------------------------------------------------------

## Summary

### Key Points to Remember:

**Joins**: - Use `inner_join()` when you only want complete matches - Use `left_join()` when you want to keep all rows from your main dataset - Use `anti_join()` when you want to find what's missing or unmatched - Always specify the `by =` argument to be clear about your join key

**Pivots**: - Use `pivot_longer()` when you have multiple columns of the same type of data - Use `pivot_wider()` when you want to compare values side by side - Long format is usually better for analysis and plotting

### Next Steps:

-   Practice with your own datasets
-   Try combining joins and pivots in the same analysis
-   Experiment with multiple join keys: `by = c("col1", "col2")`
